---
title: "Analysis the leading causes of death in Alberta"
subtitle: "Reproduce the comparison the Poisson with negative binomial models from 'Telling Stories with Data' [@alexander_2023]."
author: "Yunshu Zhang"
thanks: "Code and data are available at: https://github.com/Yunshu921/mortality_in_Alberta.git."
format: pdf
editor: visual
date: today
date-format: long
abstract: "This paper uses a dataset from the government of Alberta to fit Poisson and negative binomial models. When we focused on the top-fifteen causes of death in 2022, the result showed the negative binomial model is a better fit than the Poisson. This analysis revealed the fact that the negative binomial model may fit better than Poisson model in some reality circumstances. In addition, from the data about the cause of death, we can find the most widespread causes of death in Alberta. These insights can guide public health scientists and the policymaker in publishing healthy handbook or guidelines to decrease the mortality."
toc: TRUE
number-sections: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = "h", out.extra = "")

# Load libraries
library(tidyverse)
library(ggplot2)
library(dplyr)
library(rstanarm)
library(modelsummary)
library(broom)
library(broom.mixed)

# Read in data
alberta_causes_death <- read_csv(here::here("inputs/data/deaths_leading_causes_cleaned.csv"))

```

# Introduction

Due to the development of economy, the progress of medical and the electronization of life, our life expectancy has remarkably increased. We can perceive this point by the higher risk of developing chronic diseases compared to the past [@cite2016]. An aging population is a problem endemic to the Western and industrialized countries. Therefore, by analyzing data about causes of death, we can provide insights to improve the old-age survival. More importantly, normal people can take the information as one's heath guideline. To be more specifically, we can take aged parents to do health check for checking specific and prevalent diseases.

In this paper, we will firstly examine the relationship between the cause and the number of death for several leading causes of death by using linear regression analysis. Then we will utilize data from leading causes of death in 2022 to fit two models which are Poisson model and the negative binomial model. The estimand is which model fit better in this situation. The result of analysis showed the negative binomial model fit better than Poisson model.

After reviewing literature, they have pointed out the difference between Poisson regression and the negative binomial regression. Thus there is a gap about a specific example which shows advantages of the negative binomial model overweight that of Poisson model. Furthermore, during analying the data, some interesting findings can assist healthcare professionals and public health related department to allocate resources effectively.

In this paper, there are 4 sections, excludes the introduction. In the first section, we review the source of data from the government of Alberta, the advantages and disadvantages of data, methodologies that follow it, and data terminology. In addition, we have some plots to show the distribution of the cause and the number of death for several leading causes of death. For the second section, we run two linear regression models and explain each variable in detail. In next section, we will display the results by using tables and plots. In the final section, we discuss our results and point out some weaknesses.

This paper is carried out using statistical programming language `R` [@citeR], the library `tidyverse` [@tid] for preparation and analysis of data, the `janitor` [@jan], the `here` [@here] for read data, the `ggplot2` for generating figures, the `dplyr` [@dp], and we run the model in `R` [@citeR] using the `rstanarm` package [@rstanarm], the `broom` package [@broom], and the `broom.Mixed` package [@bm]. Since this paper is a reproduction of 'Telling Stories with Data' [@alexander_2023], some similar code were used.

# Data {#sec-data}

## Data Description and Methodology

The dataset used in this paper is from the open data of government of Alberta and can be freely downloaded at their website [@2022]. This dataset was created at 2015 May 13 and last modification was done at 2023 Sept 22. The update frequency is annual and the publisher is Service Alberta. We use this dataset not another dataset of other province since the representation of data in this dataset is clear and concise. There are some similar datasets, but the one we used is the more suitable for testing Poisson and the negative binomial models. At the same time, the government of Alberta has collected data regarding economy and finance, society and communities, employment and labour, environment, health and wellness, government, agriculture and other varied topics. The government of Alberta keeps records on a weekly,monthly, and yearly data.

The dataset consist first 30 leading causes of death from 2001 to 2022 and each cause has the corresponding number of deaths. In summary, there are 666 rows which contain four rows of descriptive words and two repeated ranking rows at row 34 and row 35. That is there are 664 variables in original dataset which is in form of csv document.

## Data Visualization

Now, we can explore this dataset by using `ggplot2` package [@gg] for generating plots, and the `knitr` package [@kn] for presenting tables.

```{r}
#| echo: false
#| warning: false
#| label: tbl-data
#| tbl-cap: "A summary table of cleaned data"

library(knitr)
kable(
  head(alberta_causes_death, 15),
  col.names = c("Year", "Cause", "Ranking", "Deaths", "Years"),
  align = c("l", "r", "r", "r", "r"),
  digits = 0,
  booktabs = TRUE,
  linesep = ""
)


```

@tbl-data shows the detail of the cleaned dataset which includes 5 variables and 663 observations in total. The variables in the dataset include Year (in years), Cause, Ranking, Death, and Years (the frequency of each cause).

```{r}
#| echo: false
#| warning: false
#| label: tbl-data-2
#| tbl-cap: "A summary table of top fifteen causes of death"

 alberta_causes_death |>
  filter(
    calendar_year == 2022,
    ranking <= 15
  ) |>
  mutate(total_deaths = format(total_deaths, big.mark = ",")) |>
   kable(
    col.names = c("Year", "Cause", "Ranking", "Deaths", "Years"),
    align = c("l", "r", "r", "r", "r"),
    digits = 0, booktabs = TRUE, linesep = ""
  )
```

@tbl-data-2 shows top-15 causes of death in Alberta in 2022 and we found some interesting things such as some diseases just showed recently, for example the COVID-19.

```{r}
#| echo: false
#| warning: false
#| label: tbl-data-3
#| tbl-cap: "A summary table of leading causes of death"

library(knitr)

alberta_causes_death_top_four <-
   alberta_causes_death |>
  filter(
    calendar_year == 2022,
    n == 22
  ) |>
  slice_max(order_by = desc(ranking), n = 4) |>
  pull(cause)

alberta_causes_death <-
  alberta_causes_death |>
  filter(cause %in% alberta_causes_death_top_four)

kable(
  head(alberta_causes_death, 15),
  col.names = c("Year", "Cause", "Ranking", "Deaths", "Years"),
  align = c("l", "r", "r", "r", "r"),
  digits = 0,
  booktabs = TRUE,
  linesep = ""
)

```

@tbl-data-3 shows some causes with variable 'Years' is equal to 22 since 2001. Firstly, we look at top-four causes of death in 2022. Since we do not want to consider some diseases get prevalent recently, these top-four causes of death have the variable 'Years' is equal to 22. Then we want to examine if there is a relationship between total deaths and each of this specific top-four causes since 2001.

```{r}
#| label: fig-1
#| fig-cap: Annual number of deaths for the top-four causes in 2022, since 2001, for Alberta, Canada
#| echo: false
#| warning: false

library(ggplot2)

alberta_causes_death |>
  ggplot(aes(x = calendar_year, y = total_deaths, color = cause)) +
  geom_line() +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Year", y = "Annual number of deaths in Alberta") +
  facet_wrap(vars(cause), dir = "v", ncol = 1) +
  theme(legend.position = "none")

```

@fig-1 shows the linear regression between different cause and time. Organic dementia and all other forms of chronic ischemic heart disease are top 1 and top2 cases of death in 2022. At the same time, malignant neoplasms of trachea, bronchus and lung and acute myocardial infarction are two causes of death at top 5 and top 6 in 2022. For the top 3 and top 4 cases of death in 2022, they do not satisfy the condition that the variable 'Years' is equal to the 22. Thus they were not in so-called top-four causes of death. When we observe these four causes, we can find that only organic dementia has showed surprising increase through 22 year. The initial deaths in 2001 was 280 and the latest death in 2022 was 2874. With the worsening aging issue, the number of deaths due to organic dementia has increased tenfold over the period of 22 years. This cause of death has also risen from the 14th position in 2001 to the current top position. In 2017, organic dementia got the top-one cause of death at the first time and then so far.

```{r}
#| echo: false
#| warning: false
#| label: tbl-data-4
#| tbl-cap: "Summary statistics of the number of yearly deaths, by cause, in 2022, in Alberta, Canada "
library(knitr)
library(dplyr)

# Calculate summary statistics by cause
summary_stats <- alberta_causes_death %>%
  summarize(
    Min = min(total_deaths),
    Mean = mean(total_deaths),
    Max = max(total_deaths),
    SD = sd(total_deaths),
    Var = var(total_deaths),
    N = n()
  )


# Define the column names and alignment
col_names <- c("Min Deaths", "Mean Deaths", "Max Deaths", "SD", "Var", "N")
align <- c("r", "r", "r", "r", "r", "r")

# Format the summary statistics as a data frame
summary_stats_df <- data.frame(
  `Min Deaths` = summary_stats$Min,
  `Mean Deaths` = summary_stats$Mean,
  `Max Deaths` = summary_stats$Max,
  `SD` = summary_stats$SD,
  `Var` = summary_stats$Var,
  `N` = summary_stats$N
)

# Print the table using kable
kable(
  summary_stats_df,
  col.names = col_names,
  align = align,
  digits = 0,
  booktabs = TRUE,
  linesep = ""
)



```

@tbl-data-4 shows the summary of statistic of total deaths in Alberta. In this table, we notice that the mean, 437, is different to the variance, 190,696.

# Model

Now we can use the `rstanarm` package [@rstanarm] to fit two models: Poisson and the negative binomial models.

```{r}
#| echo: false
#| warning: false
#| results: false


alberta_poisson <-
  stan_glm(
    total_deaths ~ cause,
    data = alberta_causes_death,
    family = poisson(link = "log"),
    seed = 996
  )

alberta_neg_binomial <-
  stan_glm(
    total_deaths ~ cause,
    data = alberta_causes_death,
    family = neg_binomial_2(link = "log"),
    seed = 996
  )

```

## Model summary of Poisson and the negative binomial

```{r}
#| echo: false
#| warning: false
#| label: tbl-model-1
#| tbl-cap: Modeling the most prevalent cause of deaths in Alberta, 2001-2021


modelsummary(
  list(
    "Poisson" = alberta_poisson,
    "Negative binomial" = alberta_neg_binomial
  )
)
```

@tbl-model-1 shows estimates for Poisson and the negative binomial models are similar. Thus we need to use other method to compare these two.

\newpage

## Model equation

Firstly, we can get the Poisson probability mass function which is same as the negative binomial is [@pitman_1993] 

$$ 
P(X=k)= e^{-λ}λ^k/k!, for\ k = 0,1,2,3,...
$$

Following an exploratory analysis of the dataset, the ultimate model is presented below.

$$ 
y_{i}|λ_{i} \sim Poisson
$$
$$
log_{e}(λ_{i}) = β_{0} + β_{1} \times cause_{i}
$$

where:

-   $y_{i}|\lambda_{i}$ is the number of deaths for different causes.
-   $\lambda_{i}$ is the intercept which means the basic mortality without any causes we control.
-   $\beta_0$ is the intercept or constant term, which represents the expected value of basic mortality when there is no active cause that we interested.
-   $\beta_1$ is the coefficient or the estimated change in mortality for a change in the cause.
-   $cause_{i}$ is the cause of the death.

In specific, we have two equations of mortality for two models:

$$ 
y_{i} = e^{log_{e}(λ_{i})}
$$
$$
y_{i} = e^{log_{e}(λ_{i})} = e^{7.037 + 0.446I_{1} + 0.224I_{2} + 0.046I_{3}}, for \ Poisson \ Model
$$
$$
y_{i} = e^{log_{e}(λ_{i})} = e^{7.039 + 0.446I_{1} + 0.221I_{2} + 0.043I_{3}}, for \ Negative \ Binomial\ Model
$$ 

where

-   $I_{1}$ means the indicator function such that if this cause happened, that is all other forms of chronic ischemic heart disease, then we get the function value of 1.Otherwise, we get the function value of 0

-   $I_{2}$ represents the indicator function such that if this cause happened, that is malignant neoplasms of trachea, bronchus and lung, then we get the function value of 1.Otherwise, we get the function value of 0

-   $I_{3}$ is the indicator function such that if this cause happened, that is organic dementia, then we get the function value of 1.Otherwise, we get the function value of 0

# Results

## Comparing posterior predictive checks for Poissson model and Negative binomial model

Since two equations have similar estimates in each row and from 'Telling Stories with Data' [@alexander_2023], we can use function `pp_check` in the `rstanarm` package [@rstanarm] to compare these two models. This involves comparing the observed outcome variable with simulations generated from the posterior distribution.

```{r}
#| label: fig-2
#| fig-cap: Poisson model
#| echo: false
#| warning: false

pp_check(alberta_poisson) +
  theme(legend.position = "bottom")

```

```{r}
#| label: fig-3
#| fig-cap: Negative binomial model.
#| echo: false
#| warning: false
pp_check(alberta_neg_binomial) +
  theme(legend.position = "bottom")

```

@fig-2 and @fig-3 are fitted plots of the Poisson model and Negative binomial model, where the dark and bold lines indicate that the negative binomial approach is a better choice for our circumstance.

Lastly, from 'Telling Stories with Data' [@alexander_2023], we can assess and compare models using the leave-one-out (LOO) cross-validation (CV) resampling method. This approach is a form of cross-validation where each fold consists of only one observation.

```{r}
#| echo: false
#| warning: false
#| label: tbl-data-5
#| tbl-cap: Compared data frame

poisson <- loo(alberta_poisson, cores = 2)
neg_binomial <- loo(alberta_neg_binomial, cores = 2)

loo_compare(poisson, neg_binomial)

```

@tbl-data-5 shows the result of comparison. Since the Negative binomial has higher value in ELPD than Poisson has, the Negative binomial model is our optimal model.

\newpage

# Discussion

## General prosess of this paper

In section three, namely the model section, we firstly examine the formula of Poisson regression. Based on the result from the data visualization, we can know that the SD and Variance of total death we want to do research is not equal. This means Poisson may not the best choice for this situation. Therefore we introduce the negative binomial regression. In the next section, we used two models to fit same dataset. At the beginning of the thing, @table-model-1 showed almost identical statistic. After applying posterior predictive checks and the leave-one-out (LOO) cross-validation (CV) resampling method, we get @fig-2, @fig-3, and @tbl-data-5. From two fit lines, we can clear infer that the negative binomial model is our optimal solution for this dataset.

## Organic dementia has taking top one cause of death

From the @fig-1, we can find that the cause of death: organic dementia has irresistibly risen to the top cause of death. After exploring the literature around the word, we find two main reason behind this. The first reason is the spectacular longevity gains. As people's lifespans extend, the likelihood of developing senile dementia also increases. And I think may not the most important reason is that dementia can easily be triggered by other illnesses. For example, from the article published at The Lancet Neurology [@2005], having a stroke will increase the risk of dementia greatly.

## Some inspiration about the leading cause of death

In the real life, data analysis about leading cause of death helps reveals the secret of different disease and shows the truth gradually. From the article: Insight on ‘typical’ longevity: An analysis of the modal lifespan by leading causes of death in Canada [@cite2016], the greatest gap in modal age-at-death estimates is seen between lung cancer and cardiovascular diseases. When we look at the long time range, from 1974 to 2011, for male, there was a 10 year gap between lung cancer and cardiovascular diseases. For female, there was a nearly 15 years gap between lung cancer and cardiovascular diseases. These two gaps maintained for almost 40 years. Although the modal age is postpones, the difference of two cause of death doesn't change. From this point, doctors in hospital can make more targeted judgments about different diseases.

## Weaknesses

During constructing the model, we can find the @tbl-model-1 only has three predictors not four. In this part, we need to examine the principle behind the code. I think give a more detail explanation should be a better solution. In addition, for the overall logic of this paper, we need do more work. For instance, we should use concise and simple sentence to interpret relative complex definitions.

## Improvement

We are supposed to do more examples related to comparing different models. Because in this way we can learn thing better. At the same time, we need read and pracice all the time to keep our mind clear and positive. For this paper, I think the detail of processing data and the wording can be improved.

\newpage

# References
